Building a continuous benchmarking ecosystem in bioinformatics

### 2025-02-10 review_cycle_1/highlighted_manifesto_main.tex
syntax fixes

Benchmarking (defined below) is a critical step in method development. Such studies are typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{cao2023-jz}. A benchmarking study can be run by a single person or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

In our discussion, we outline concepts and specifications for an ideal benchmarking system; we examine philosophical issues like gatekeeping and trust; and provide some technical recommendations, such as standardization.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig_benchmarking.pdf}
    {\color{red} \caption[Run types, properties, capabilities and features.]{Solo- and community-benchmarks, workflows, and benchmarking system-enabled (\textit{bs}) benchmarks differ in run properties and their consequences in terms of capabilities and features, including whether: inputs, truths, hidden truths, metrics or run results are available for readers; code is free software, software environments are managed, and execution automated; there is a clear track of (human) contributors; results are run in comparable hardware; an entrypoint to results or result interpretation are provided. These run intrinsic properties have consequences in terms of features (i.e. transparency, forkability, availability for metaanalysis) and technical capabilities (i.e. software versioning or ability to run anonymously and locally (stealth mode). \label{fig:definitions}}
\end{figure}

@@

Can benchmarking be systematized? {\color{red}Benchmarking aims to calculate, collect and report performance metrics of methods aiming to solve a task. Such process normally requires collecting the data, running the methods in an automated manner. Some of these tasks can be automated by workflow systems. Others, such as sharing results in an interactive manner and, particularly, result interpretation, go beyond what workflows offer. Similarly, collaborative benchmarking and challenges add extra layers of complexity, including collaboration tracking, hidding ground truths, or the ability to release multiple versions of the same benchmark as it grows (Figure~\ref{fig:definitions}). Moreover, scientific benchmarking borrows procedures from the scientific method and from scientific publishing, hence having expectations of fairness, reproducibility, transparency, and trust. 
\sout{Benchmarks should be orchestrated and run by}} \emph{Benchmarking systems} {\color{red} aim to orchestrate the workflow management, benchmarking and community engagement aspects of the process  \sout{to create} to generate benchmark} `artifacts' (code snapshots, file outputs to be shared, performance outputs) {\color{red} in a systematic way and following good practices and standards. Namely,}  the various elements of code, especially the details on how datasets are preprocessed and how methods are run, should be explicitly available for scrutiny. Such {\color{red}\sout{a}benchmarking} systems should facilitate components to be public, open and allow contributions (e.g., corrections or new components), such that the benchmarks of the future can run more continuously. At the outset of a benchmark, a benchmark `definition' could be envisioned, which could give a formal specification of the entire set of components (and the pattern of artifacts to be generated). The benchmark definition could even be expressed as a single configuration file that specifies the scope and topology of components to include, details of the repositories (code implementations with versions), the instructions to create software environments (accessible across compute architectures), the parameters used, and what components to snapshot for a release. A benchmark definition in terms of layers and their corresponding challenges and opportunities within each layer is illustrated in Figure~\ref{fig:layers}.}

@@

{\color{red} Discuss DREAM, CASP etc and scientific communities here (reviewer 1 and 2).}

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning appropriate hardware; handling a reproducible and efficient software stack; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results {\color{red}(Figure~\ref{fig:definitions})}.

@@

\subsection*{Reproducible software environments}

@@

\item Software dependencies cause the biggest impact on replicability and maintainability; archiving all combinations of a large dependency tree, across several OS images, exponentially increases retention and maintenance costs. Making the long-term replicability of a benchmark tractable implies adopting a sane software management system. Several solutions exist, from containerization (e.g., apptainer or singularity) \cite{Kurtzer2017-mn} and automation of reproducible and efficient software management (\texttt{EasyBuild} \cite{Hoste2012-gg} or \texttt{Spack} \cite{Gamblin2015-ll}) to distributing software installations such as CernVM-FS \cite{Blomer2013-jk},  or combinations of the two such as the Digital Research Alliance of Canada (former \texttt{computecanada}) \cite{Boissonneault2019-wm} and \texttt{EESSI} \cite{Droge2023-ax}.

@@

Similarly, even if all benchmark code were free software, running it could be hindered by (non-free) proprietary interpreters (e.g., MATLAB). Benchmarking systems should be explicit in allowing or disallowing proprietary software, and be aware and track EULAs (end user license agreements). This conundrum extends to the operating system layer and to system dependencies (e.g., compilers). In terms of benchmark re-runnability, fully free software system dependencies arguably offer the highest reusability, and can be handled in a reproducible manner across operating systems and system architectures (e.g., amd64, arm64) \cite{Droge2023-ax}.

>>>>

Benchmarking (defined below) is a critical step in method development. Such studies are typically published as either `methods-development papers' (MDPs) where new methods are compared against existing ones; or `benchmark-only papers' (BOPs), where a set of existing methods are compared in a more neutral way, as defined in \cite{cao2023-jz}. A benchmarking study can be run by a single person {\color{red} (solo benchmarking)} or a small group, or it can be organized as a `challenge' or within a hackathon. In this work, we present a perspective on building a robust benchmarking ecosystem in bioinformatics, coming primarily from the methodologist's viewpoint (researchers developing computational tools).  While our focus is bioinformatics, we aim to abstract principles and connect to similar challenges in adjacent fields.

In our discussion, we outline concepts {\color{red} related to workflow automation, benchmark formalization and benchmarking-specific tasks beyond the former, including \sout{and specifications for an ideal benchmarking system; we examine}} philosophical issues like gatekeeping and trust; and provide some technical recommendations, such as standardization.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/fig_benchmarking.pdf}
    {\color{red} \caption[Run types, properties, capabilities and features.]{Solo- and community-benchmarks, workflows, and benchmarking system-enabled (\textit{bs}) benchmarks differ in run properties and their consequences in terms of capabilities and features, including whether: inputs, truths, hidden truths, metrics or run results are available for readers; code is free software, software environments are managed, and execution automated; there is a clear track of (human) contributors; results are run in comparable hardware; an entrypoint to results or result interpretation are provided. These run intrinsic properties have consequences in terms of features (i.e. transparency, forkability, availability for metaanalysis) and technical capabilities (i.e. software versioning or ability to run anonymously and locally (stealth mode). \label{fig:definitions}}
    }
\end{figure}

@@

Can benchmarking be systematized? {\color{red}Benchmarking aims to calculate, collect and report performance metrics of methods aiming to solve a task. Such process normally requires collecting appropriate data and running methods in an automated manner. Some of these tasks can be automated by workflow systems. Others, such as sharing results in an interactive manner and, particularly, result interpretation and making results reusable by others, go beyond what workflows offer. Similarly, collaborative benchmarking and challenges add extra layers of complexity, including collaboration tracking, hiding ground truths, or the ability to release multiple versions of the same benchmark as it develops (Figure~\ref{fig:definitions}). Moreover, scientific benchmarking borrows procedures from the scientific method and from scientific publishing, hence having expectations of fairness, reproducibility, transparency, and trust. 
\sout{Benchmarks should be orchestrated and run by}} \emph{Benchmarking systems} {\color{red} aim to orchestrate the workflow management, benchmarking and community engagement aspects of the process  \sout{to create} to generate benchmark} `artifacts' (code snapshots, file outputs to be shared, performance outputs) {\color{red} in a systematic way and following good practices and standards. Namely,}  the various elements of code, especially the details on how datasets are preprocessed and how methods are run, should be explicitly available for scrutiny. Such {\color{red}\sout{a}benchmarking} systems should facilitate components to be public, open and allow contributions (e.g., corrections or new components), such that the benchmarks of the future can run more continuously. At the outset of a benchmark, a benchmark `definition' could be envisioned, which could give a formal specification of the entire set of components (and the pattern of artifacts to be generated). The benchmark definition could even be expressed as a single configuration file that specifies the scope and topology of components to include, details of the repositories (code implementations with versions), the instructions to create software environments (accessible across compute architectures), the parameters used, and what components to snapshot for a release. A benchmark definition in terms of layers and their corresponding challenges and opportunities within each layer is illustrated in Figure~\ref{fig:layers}.

@@

{\color{red} Collaborative benchmarking efforts enunciate field-specific challenges and provide a venue for \emph{communities} to address them. For instance, DREAM

GIAB

CASP

have successfully shown how collaborative benchmarks can help in moving fields forward.}

@@

Benchmarking as a process deals with tasks other than workflow definition and execution, including: collecting and keeping track of contributions from users; provisioning appropriate hardware; handling a reproducible and efficient software stack {\color{red}(the advantages of decoupling environments handling from workflow execution is further discussed in section~\ref{softwareenvs})}; managing storage and access-control lists; rendering results dashboards and summarizing results; versioning code, workflow runs, and files; and providing extensive documentation from defining scope or contribution guidelines, to transparent logging and crediting contributions and disseminating results {\color{red}(Figure~\ref{fig:definitions})}.

@@

\subsection*{Reproducible software environments \label{softwareenvs}}

@@

\item Software dependencies cause the biggest impact on replicability and maintainability; archiving all combinations of a large dependency tree, across several OS images, exponentially increases retention and maintenance costs. Making the long-term replicability of a benchmark tractable implies adopting a sane software management system. Several solutions exist, from containerization (e.g., apptainer or singularity) \cite{Kurtzer2017-mn} and automation of reproducible and efficient software management (\texttt{EasyBuild} \cite{Hoste2012-gg} or \texttt{Spack} \cite{Gamblin2015-ll}) to distributing software installations such as CernVM-FS \cite{Blomer2013-jk},  or combinations of the two such as the Digital Research Alliance of Canada (former \texttt{computecanada}) \cite{Boissonneault2019-wm} and \texttt{EESSI} \cite{droge2023-ax}.

@@

Similarly, even if all benchmark code were free software, running it could be hindered by (non-free) proprietary interpreters (e.g., MATLAB). Benchmarking systems should be explicit in allowing or disallowing proprietary software, and be aware and track EULAs (end user license agreements). This conundrum extends to the operating system layer and to system dependencies (e.g., compilers). In terms of benchmark re-runnability, fully free software system dependencies arguably offer the highest reusability, and can be handled in a reproducible manner across operating systems and system architectures (e.g., amd64, arm64) \cite{droge2023-ax}.
